Non-word repetition in Yélî Dnye

### 2021-02-09 manuscript.Rmd
incorporated mcs changes

abstract: >
   In nonword repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in their language, and asked to repeat this item as closely as possible. NWR scores are thought to reflect some aspects of phonological development, saliently a perception-production loop supporting flexible production patterns. In this study, we report on NWR results among children learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea. Our overarching goal is to reflect on how NWR scores can be compared across participants, studies, languages, and populations, in order to shed light on the factors universally structuring variation  in language development. More specifically, this study contributes to three lines of research. First, we contribute to investigations on NWR across diverse languages, by documenting that, in Yélî Dnye, non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects. Second, contributing to mounting research suggesting that length effects may be language- or population-specific, we find rather weak effects of item length. Third, we add a datapoint on potential sources of individual variation effects, by establishing that in our sample age has a strong effect on NWR scores, whereas there are weak correlations with gender, maternal education, and birth order.  Together, these data provide a unique view of online phonological processing in an understudied language while making preliminary connections between language development and cross-linguistic features.

keywords          : "phonology, non-word repetition, development"
wordcount         : "~9,000 words"

@@

## TODO Alex
- Shorten the abstract a little please (it's very clearly written now but probably too long)
- Other keywords for AC to consider adding: Papuan, non-industrial/non-urban, comparative, typology/markedness
- Changed "Scores" to "scores" in the figures—check if I missed anything!

## TODO Middy

- read over whole thing -- does the logic sound ok?
- tell me if you think you CAN'T LIVE with my turning this in with  figures 1 and 5 as they are (I promise I'll improve them in the revision stage)

@@

The first research area is at the crossing of typology and phonological development.  Previous work using NWR has preferred relatively universal and early-acquired phonemes [with the possible exception of @gallagher2014identity], in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects [@gallon2007non] and in part because the test is sometimes used to measure working memory in the context of executive functions [@mulder2017early] rather than purely linguistic skill <!--Previous sentence change: "language" >> "linguistic skill-->. Here, we investigate repetition of non-word items containing cross-linguistically common and cross-linguistically rare phonetic targets. Specifically, we included a subset of non-word items with typologically rare sounds to ask whether these sounds are disadvantaged in the perception-production loop involved in NWR.

Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development within specific languages. Some work documents much lower NWR scores for longer, compared to shorter, items [e.g., among Cantonese-learning children; @stokes2006nonword], whereas differences are negligible in other studies  [e.g., among Italian learners; @piazzalunga2019articulatory]. It is possible that differences are due to language characteristics, including the modal length of words in the language and/or in child-directed speech in that culture. In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words are modally longer. To see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths. We found data for learners of Israeli Arabic  [@jabere2018xperiment]<!--Previous sentence change: added square brackets around jaber-awida ref-->; Cantonese  [@stokes2006nonword]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant]; and integrated those data with Yélî Dnye results from the present study in Figure \@ref(fig:fig-prevlit).

@@

Our reading of this Figure is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically line <!--Changed previous bit: "lign" >> "line"--> up with expected word length in different languages. For instance, the difference in NWR scores for 2- versus 3-syllable items (averaging across age groups) is largest in Tsimane' (~`r round(difs[difs$Group.1=="Tsimane'","x"])`%) and Arabic (~`r round(difs[difs$Group.1=="Arabic","x"])`%), which tend to have longer words, as does Italian, where the difference between 2- and 3-syllable items was only ~`r round(difs[difs$Group.1=="Italian","x"])`%. Similarly, two languages that are often described as heavily biased towards monosyllables show diverse length effects (Cantonese ~`r round(difs[difs$Group.1=="Cantonese","x"])`% versus English ~`r round(difs[difs$Group.1=="English","x"])`%). Given the paucity of research looking at this question, and the diversity of current results, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic analyses <!--Changed previous bit: "approaches" >> "analyses"-->.

@@

Fourth, these data contribute to the small literature using this task with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory [see @maddieson2005correlating for a broad classification of languages based on inventory size]. Indeed, NWR has seldom been used outside of urban settings in Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant], nor with languages having large phonological inventories [e.g., more than 34 consonants and 7 vowel qualities @maddieson2013vowel and @maddieson2013consonant with no exceptions to our knowledge] <!--Changed previous bit to fix the paren display w/ the refs-->. There are no theoretical reasons to presume that the technique will not generalize to these new conditions. That said, @cristia2020infant recently reported relatively lower NWR scores among the Tsimane', a non-Western rural population, interpreting these findings as consistent with the hypothesis that lower levels of infant-directed speech and/or low prevalence of literacy in a population could lead to population-level differences in NWR scores. In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, for instance better than rhyme awareness [e.g., @gathercole1991differentiating]. Therefore, it may not be a general index of phonological development, but more specifically reflect certain non-universal skills. Indeed, @cristia2020infant present the task as being a good index of the development of "short-hand-like" representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations. To our knowledge, there is little discussion of linguistic effects -- i.e., of potential differences in NWR as a function of language typology -- or cultural effects -- i.e., of potential differences in NWR as a function of other differences across human populations, aside from @cristia2020infant's hypotheses just mentioned. Regarding potential language differences, we note that the very fact that studies compose items by varying syllable structure and word length, while prefering relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/) may indicate a bias towards Indo-European languages, where syllable structure and word length are indeed important structural dimensions. This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first people using the method, who tend to work on English.

Before going into the details of our study design we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island. As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites, which are more generally accessible <!--Changed previous bit: "typically easily" >> "more generally"-->. Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from Rossel Island is both infrequent and irregular. International phone calls and digital exchanges that require significant data transfer are typically not an option. Data collection is therefore typically limited to the duration of the researchers' on-island visits.

@@

Yélî Dnye is an isolate language (presumed Papuan) spoken by approximately 7,000 people residing on Rossel Island, an island found at the far end of the Louisiade Archipelago in Milne Bay Province, Papua New Guinea. The Yélî sound system, much like its baroque grammatical system [@levinsonYDgrammar], is unlike any other in the region. In total, Yélî Dnye uses 90 distinctive segments (not including an additional three rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages [@foley1986papuan; @levinsonYDgrammar; @maddiesonIPphoneticsYD]. Thus, with respect to our first research goal, Yélî Dnye is <!--Changed previous bit: "seemed is" >> "is"-->a good language to attempt an investigation on NWR with sounds varying in cross-linguistic frequency because of its large inventory, which includes some rare sounds.

@@

Our knowledge of Yélî language development is growing [e.g., @brown2011cultural; @brown2014interactional; @brownIPchildrearing; @casillas2020early; @liszkowski2012prelinguistic], but research into Yélî phonological development has only just begun. For example, Peute and colleagues' [-@peuteIPconsonants] find that Yélî Dnye-learning children's early spontaneous consonant productions appear to exclusively feature simplex and typologically frequent phones<!--Changed previous bit: added what I think the relevant bit from Bram's paper is -- change as needed!-->. We hope the present study contributes to this growing line of work<!--Changed previous bit: "filling this gap" >> "this growing line of work"-->.

@@

Some aspects of the community are relevant for contextualizing our study design and results, particularly regarding sources of individual variation<!--Changed previous sentence: new wording-->. Specifically, we investigated potential effects of age, gender, maternal education, and birth order. There is nothing particular to note regarding age and gender, but we have some comments that pertain to the other two factors.

The typical household in our dataset includes seven individuals (typically, a mixed sex couple and children---their own and possibly some<!--Minor wording change: " -- their own and " >> "--their own and possibly some"--> billeting others, as discussed in the next paragraph) and is situated among a collection of four or more other households, with structures often arranged around an open grassy area. These household clusters are organized by patrilocal relation, such that they typically comprise a set of brothers, their wives and children, and their mother and father, with neighboring hamlets also typically related through the patriline. Land attribution for building one's home is decided collectively based on land availability<!--I removed this bit since I think there's some more nuance involved here and it seems better to not get into it: ", and typically does not take into consideration an individual's desire to be close to a school. "-->

Most Yélî parents are swidden horticulturalists<!--I removed this bit because I don't know what you mean: , and those who are not may not reside in the island.--> Within a group of households, it is often the case that <!--removed "most" here-->older adolescent and adults spend their day tending to their gardens (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities, which leave them little time to spend directly with the children in their household (other than infants).<!--Mmmmm this is not true. There's a great deal of socializing time by men and women during the day that isn't centered on some kind of productive domestic activity (maybe it'd be too far to call it leisure time, but it's not that off-the-mark!) The description you give suggests that they're too busy to hang out with the kids and that isn't an accurate characterization of what's going on. It's also the case that kids start doing significant and fairly independent domestic work before older adolescence, especially if they're not in school at the moment. I think, e.g., both Yaawaa and Taakeme had their own starter gardens to manage before age 10 and were expected to contribute to family dinner from their bounty :)--> Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 members) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2020early]. Formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond of what is locally available [@casillas2020early]. Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to billet their school-aged relatives during the weekdays for long segments of the school year. Children start school often at around age six or seven <!--added "or seven"-->, although the precise age depends on the child's readiness, as judged by their teacher<!--replaced "apparent development" with "readiness, as judged by their teacher"-->.

Some general ideas regarding potential maternal education effects on our data may be drawn from the observations above. To begin with, many of our participants above 6 years of age may not be living with their birth mother but with other relatives, which may weaken maternal education effects. Additionally, the importance given to formal education appears relatively stable over the period that Rossel Island has been visited by language researchers (Steven Levinson and Penelope Brown, about 20 years). Together with the fact that land attribution for hamlet development is not made on the basis of proximity to a school<!--changed the wording of the prior sentence-->, it seems to us that the length of formal education a given individual may have is not necessarily a good index of their socio-economic status or other individual properties, unlike what happens in industrialized sites, and variation may simply due to random factors like living close to a school or having relatives there.

As for birth order, much of the work on birth order effects on cognitive development (including language) has been carried out in the last 70 years and in agrarian or industrialized settings [@gratz2018competition;@barclay2015within], where nuclear families are more likely to be the prevalent rearing environment [@lancy2015anthropology]. It is possible that birth order effects are stronger in such a setting, because much of the stimulation can only come from the parents, and when there are multiple children, the inter-birth interval is small enough that older siblings may not be of an age that allows them to contribute to their younger siblings' stimulation. This contrasts with this picture just drawn in the Yélî community, where children will typically benefit from a rich and extensive socially stimulating setting, surrounded by siblings, and cousins of several orders, regardless of their birth order in their nuclear family<!--moved the "regardless of their birth order in their nuclear family" phrase to the end-->.

We add some observations that will help us integrate this study to the broader investigation of NWR across cultures. As mentioned previously, there is one report of lower NWR scores among the Tsimane', which the authors interpret as consistent with long-term effects of low levels of infant-directed speech [@cristia2020infant]. However, @cristia2020infant also point out that this is based on between-paper comparisons, and thus methods and a myriad other factors have not been controlled for. The Yélî community can help us shed further light onto<!--minor wording change to "light" metaphor--> this question because direct speech to children under 3;0 is relatively infrequent in this community too [@casillas2020early]. Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggests that Tsimane' young children are spoken to about 4.2 minutes per hour [@scaff2021daylong], and Yélî children about 3.6 minutes per hour [@casillas2020early]. Thus, if input quantities in early childhood are a major determinant of NWR scores, we should observe similarly low NWR scores as in @cristia2020infant.

@@

A first list of candidate items was generated during a trip to the island in 2018 by selecting simple consonants (/p/,	/t/,	/ṭ/,	/k/,	/m/,	/n/,	/w/,	/y/) and vowels (/i/, /o/, /u/, /a/, /e/) and combining them into consonant-vowel syllables, then sampling the space of 2- to 4-syllable sequences. These candidates were automatically removed from consideration if they appeared in Levinson's (2015) dictionary. The second author presented them orally to three local research assistants, all native speakers of Yélî Dnye, who repeated each form as they would in an NWR task and additinally let the experimenter know if the item was in fact a word or phrase in Yélî Dnye. Any item reported to have a meaning or a strong association with another word form or meaning was excluded.

@@

To inform our analyses, we estimated the typological frequency of all phonological segments present in the target items using the PHOIBLE cross-linguistic phonological inventory database [@phoible]. For each phone in our task, we extracted the number and percentage of languages noted to have that phone in its inventory. While PHOIBLE is an unprecedentedly comprehensive database, with phonological inventory data for over 2000 languages at the time of writing, it is of course still far from complete, which may mean that frequencies are estimates rather than precise descriptors). Note that nearly half of the segment types are only attested in one language (Steven Moran, personal communication). Extrapolating from this observation, we treat the three segments in our stimuli that were unattested in PHOIBLE (/lβʲ/, /ṭp/, and /tp/) as having a frequency of 1 (i.e., appearing in one language), with a (rounded) percentile of 0% (i.e., its cross-linguistic percentile is zero).

@@

We tested a total of `r dim(demo)[1]` children from `r length(levels(factor(demo$familyID)))` families spread across four hamlets. We excluded test sessions from analysis for the following reasons: refused participation or failure to repeat items presented over headphones even after coaching (N=`r sum(demo$included=="didn\xeat repeat")`), spoke too softly to allow offline coding (N=`r sum(demo$included=="spoke too softly")`), or were 13 years old or older (N=`r sum(demo$Age >12 | demo$included=="too old",na.rm=T)`; we tested these teenagers to put younger children at ease). The remaining `r dim(inc)[1]` children (`r sum(inc$Sex=="G")` girls) were aged from `r min(inc$Age, na.rm=T)` to `r max(inc$Age, na.rm=T)` years (M = `r round(mean(inc$Age, na.rm=T),1)` years, SD = `r round(sd(inc$Age, na.rm=T),1)` years).  In terms of birth order, `r birthorder.tab[[1]]` were first borns, `r birthorder.tab[[2]]` second,  `r birthorder.tab[[3]]` third,  `r birthorder.tab[[4]]` forth,  `r birthorder.tab[[5]]` fifth, and `r birthorder.tab[[6]]` sixth, with birth order missing for `r sum(is.na(inc$birthOrder))` children. These children were tested in a remote hamlet, and we unfortunately did not ask about birth order before leaving the site. Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years).[^3] We also note that there were `r sum(inc$monolingual=="yes", na.rm=T)` only exposed to Yélî Dnye at home, `r sum(inc$monolingual=="no", na.rm=T)` children exposed to Yélî Dnye plus one or more other languages at home.[^2]

[^2]: Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English. While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission---the same region in which the current data were collected---where there is a higher incidence of married-in mothers from other islands [@brownIPchildrearing]. Children in these multilingual families grow up speaking Yélî Dnye plus English, Tok Pisin, and/or other language(s) from the region.

@@

Taking into account only the first attempts, we derived overall averages across all items. The overall NWR score was M = `r round(mean(overall$Word))`% (SD = `r round(sd(overall$Word))`%), Cohen's d = `r cohensd(mean(overall$Word),sd(overall$Word))`. The phoneme-based normalized Levenshtein distance was  M = `r round(mean(overall$NLD))`% (SD = `r round(sd(overall$NLD))`%), meaning that about a fifth of phonemes were substituted, inserted, or deleted.

@@

Table \@ref(tab:tab-c) showed that errors involving complex consonants targets were more common than those involving oral vowels (`r 100-mytab["Complex Target","% Corr."]` versus `r 100-mytab["Simple Target","% Corr."]`). Additionally, errors in which a complex consonant was mispronounced as a simple consonant were quite common, whereas those in which a simple consonant was produced as a complex one were vanishingly rare.

@@

### Factor structuring individual variation

@@

Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals, as per our third research question. As shown in Figure \@ref(fig:fig-scores-by-age), there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (`r corage$statistic`) = `r corage$estimate`, p `r pval.display(corage$p.value)`). In contrast, there was no clear association between NWR scores and sex (Welch t (`r tsex$parameter`) = `r tsex$statistic`, p `r pval.display(tsex$p.value)`), birth order  (data missing for `r sum(is.na(demo$birthOrder[demo$included=="yes"]))` children, rho = (`r round(corbo$statistic,1)`) = `r corbo$estimate`, p `r pval.display(corbo$p.value)`), or maternal education  (rho ( `r round(corme$statistic,1)`) = `r corme$estimate`, p `r pval.display(corme$p.value)`).

>>>>

abstract: >
   In nonword repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in their language, and asked to repeat this item as closely as possible. NWR scores are thought to reflect some aspects of phonological development, saliently a perception-production loop supporting flexible production patterns. In this study, we report on NWR results among children learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea. Our overarching goal is to reflect on how NWR scores can be compared across participants, studies, languages, and populations, in order to shed light on the factors universally structuring variation  in language development at a global and individual level via three contributions. First, we found that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects. Second, we documented rather weak effects of item length. Third, we found that age has a strong effect on NWR scores, whereas there are weak correlations with gender, maternal education, and birth order.  Together, these data provide a unique view of online phonological processing in an understudied language while making preliminary connections between language development and cross-linguistic features.

keywords          : "phonology, non-word repetition, development, Papuan, non-industrial, non-urban, comparative, typology, markedness"
wordcount         : "~11,00 words"

@@

@@

The first research area is at the crossing of typology and phonological development.  Previous work using NWR has preferred relatively universal and early-acquired phonemes [with the possible exception of @gallagher2014identity], in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects [@gallon2007non] and in part because the test is sometimes used to measure working memory in the context of executive functions [@mulder2017early] rather than purely linguistic skill. Here, we investigate repetition of non-word items containing cross-linguistically common and cross-linguistically rare phonetic targets. Specifically, we included a subset of non-word items with typologically rare sounds to ask whether these sounds are disadvantaged in the perception-production loop involved in NWR.

Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development within specific languages. Some work documents much lower NWR scores for longer, compared to shorter, items [e.g., among Cantonese-learning children; @stokes2006nonword], whereas differences are negligible in other studies  [e.g., among Italian learners; @piazzalunga2019articulatory]. It is possible that differences are due to language characteristics, including the modal length of words in the language and/or in child-directed speech in that culture. In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words are modally longer. To see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths. We found data for learners of Israeli Arabic  [@jabere2018xperiment]; Cantonese  [@stokes2006nonword]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant]; and integrated those data with Yélî Dnye results from the present study in Figure \@ref(fig:fig-prevlit).

@@

Our reading of this Figure is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically line up with expected word length in different languages. For instance, the difference in NWR scores for 2- versus 3-syllable items (averaging across age groups) is largest in Tsimane' (~`r round(difs[difs$Group.1=="Tsimane'","x"])`%) and Arabic (~`r round(difs[difs$Group.1=="Arabic","x"])`%), which tend to have longer words, as does Italian, where the difference between 2- and 3-syllable items was only ~`r round(difs[difs$Group.1=="Italian","x"])`%. Similarly, two languages that are often described as heavily biased towards monosyllables show diverse length effects (Cantonese ~`r round(difs[difs$Group.1=="Cantonese","x"])`% versus English ~`r round(difs[difs$Group.1=="English","x"])`%). Given the paucity of research looking at this question, and the diversity of current results, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic analyses.

@@

Fourth, these data contribute to the small literature using this task with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory [see @maddieson2005correlating for a broad classification of languages based on inventory size]. Indeed, NWR has seldom been used outside of urban settings in Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant], nor with languages having large phonological inventories [e.g., more than 34 consonants and 7 vowel qualities @maddieson2013vowel and @maddieson2013consonant with no exceptions to our knowledge]. There are no theoretical reasons to presume that the technique will not generalize to these new conditions. That said, @cristia2020infant recently reported relatively lower NWR scores among the Tsimane', a non-Western rural population, interpreting these findings as consistent with the hypothesis that lower levels of infant-directed speech and/or low prevalence of literacy in a population could lead to population-level differences in NWR scores. In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, for instance better than rhyme awareness [e.g., @gathercole1991differentiating]. Therefore, it may not be a general index of phonological development, but more specifically reflect certain non-universal skills. Indeed, @cristia2020infant present the task as being a good index of the development of "short-hand-like" representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations. To our knowledge, there is little discussion of linguistic effects -- i.e., of potential differences in NWR as a function of language typology -- or cultural effects -- i.e., of potential differences in NWR as a function of other differences across human populations, aside from @cristia2020infant's hypotheses just mentioned. Regarding potential language differences, we note that the very fact that studies compose items by varying syllable structure and word length, while prefering relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/) may indicate a bias towards Indo-European languages, where syllable structure and word length are indeed important structural dimensions. This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first people using the method, who tend to work on English.

Before going into the details of our study design we first give an overview of Yélî Dnye phonology as well as a brief ethnographic review of the developmental environment on Rossel Island. As discussed above, NWR has been almost exclusively used in urban, industrialized populations, so we provide this additional ethographic information to contextualize the adaptations we have made in running the task and collecting the data, compared to what is typical in commonly studied sites, which are more generally accessible. Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from Rossel Island is both infrequent and irregular. International phone calls and digital exchanges that require significant data transfer are typically not an option. Data collection is therefore typically limited to the duration of the researchers' on-island visits.

@@

Yélî Dnye is an isolate language (presumed Papuan) spoken by approximately 7,000 people residing on Rossel Island, an island found at the far end of the Louisiade Archipelago in Milne Bay Province, Papua New Guinea. The Yélî sound system, much like its baroque grammatical system [@levinsonYDgrammar], is unlike any other in the region. In total, Yélî Dnye uses 90 distinctive segments (not including an additional three rarely used consonants), far outstripping the phonemic inventory size of other documented Papuan languages [@foley1986papuan; @levinsonYDgrammar; @maddiesonIPphoneticsYD]. Thus, with respect to our first research goal, Yélî Dnye is a good language to attempt an investigation on NWR with sounds varying in cross-linguistic frequency because of its large inventory, which includes some rare sounds.

@@

Our knowledge of Yélî language development is growing [e.g., @brown2011cultural; @brown2014interactional; @brownIPchildrearing; @casillas2020early; @liszkowski2012prelinguistic], but research into Yélî phonological development has only just begun. For example, Peute and colleagues' [-@peuteIPconsonants] find that Yélî Dnye-learning children's early spontaneous consonant productions appear to exclusively feature simplex and typologically frequent phones. We hope the present study contributes to this growing line of work.

@@

Some aspects of the community are relevant for contextualizing our study design and results, particularly regarding sources of individual variation. Specifically, we investigated potential effects of age, gender, maternal education, and birth order. There is nothing particular to note regarding age and gender, but we have some comments that pertain to the other two factors.

The typical household in our dataset includes seven individuals (typically, a mixed sex couple and children---their own and possibly some billeting others, as discussed in the next paragraph) and is situated among a collection of four or more other households, with structures often arranged around an open grassy area. These household clusters are organized by patrilocal relation, such that they typically comprise a set of brothers, their wives and children, and their mother and father, with neighboring hamlets also typically related through the patriline. Land attribution for building one's home is decided collectively based on land availability.

Most Yélî parents are swidden horticulturalists. Within a group of households, it is often the case that older adolescent and adults spend their day tending to their gardens (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities, which leaves them less time to spend directly with the children in their household than may be the case for stay-at-home parents in urban sites. Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (~10 members) independent and mixed-age child play groups [@brownIPchildrearing; @casillas2020early]. Formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond of what is locally available [@casillas2020early]. Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to billet their school-aged relatives during the weekdays for long segments of the school year. Children start school often at around age seven, although the precise age depends on the child's readiness, as judged by their teacher.

Some general ideas regarding potential maternal education effects on our data may be drawn from the observations above. To begin with, many of our participants above 6 years of age may not be living with their birth mother but with other relatives, which may weaken maternal education effects. Additionally, the importance given to formal education appears relatively stable over the period that Rossel Island has been visited by language researchers (Steven Levinson and Penelope Brown, about 20 years). Overall, it seems to us that the length of formal education a given individual may have is not necessarily a good index of their socio-economic status or other individual properties, unlike what happens in industrialized sites, and variation may simply due to random factors like living close to a school or having relatives there.

As for birth order, much of the work on birth order effects on cognitive development (including language) has been carried out in the last 70 years and in agrarian or industrialized settings [@gratz2018competition;@barclay2015within], where nuclear families are more likely to be the prevalent rearing environment [@lancy2015anthropology]. It is possible that birth order effects are stronger in such a setting, because much of the stimulation can only come from the parents, and when there are multiple children, the inter-birth interval is small enough that older siblings may not be of an age that allows them to contribute to their younger siblings' stimulation. This contrasts with this picture just drawn in the Yélî community, where children will typically benefit from a rich and extensive socially stimulating setting, surrounded by siblings, and cousins of several orders, regardless of their birth order in their nuclear family.

We add some observations that will help us integrate this study to the broader investigation of NWR across cultures. As mentioned previously, there is one report of lower NWR scores among the Tsimane', which the authors interpret as consistent with long-term effects of low levels of infant-directed speech [@cristia2020infant]. However, @cristia2020infant also point out that this is based on between-paper comparisons, and thus methods and a myriad other factors have not been controlled for. The Yélî community can help us shed further light onto this question because direct speech to children under 3;0 is relatively infrequent in this community too [@casillas2020early]. Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggests that Tsimane' young children are spoken to about 4.2 minutes per hour [@scaff2021daylong], and Yélî children about 3.6 minutes per hour [@casillas2020early]. Thus, if input quantities in early childhood are a major determinant of NWR scores, we should observe similarly low NWR scores as in @cristia2020infant.

@@

A first list of candidate items was generated during a trip to the island in 2018 by selecting simple consonants (/p/,	/t/,	/ṭ/,	/k/,	/m/,	/n/,	/w/,	/y/) and vowels (/i/, /o/, /u/, /a/, /e/) and combining them into consonant-vowel syllables, then sampling the space of resulting possible 2- to 4-syllable sequences. These candidates were automatically removed from consideration if they appeared in Levinson's (2015) dictionary. The second author presented them orally to three local research assistants, all native speakers of Yélî Dnye, who repeated each form as they would in an NWR task and additionally let the experimenter know if the item was in fact a word or phrase in Yélî Dnye. Any item reported to have a meaning or a strong association with another word form or meaning was excluded.

@@

To inform our analyses, we estimated the typological frequency of all phonological segments present in the target items using the PHOIBLE cross-linguistic phonological inventory database [@phoible]. For each phone in our task, we extracted the number and percentage of languages noted to have that phone in its inventory. While PHOIBLE is an unprecedentedly comprehensive database, with phonological inventory data for over 2000 languages at the time of writing, it is of course still far from complete, which may mean that frequencies are estimates rather than precise descriptors. Note that nearly half of the segment types are only attested in one language (Steven Moran, personal communication). Extrapolating from this observation, we treat the three segments in our stimuli that were unattested in PHOIBLE (/lβʲ/, /ṭp/, and /tp/) as having a frequency of 1 (i.e., appearing in one language), with a (rounded) percentile of 0% (i.e., its cross-linguistic percentile is zero).

@@

We tested a total of `r dim(demo)[1]` children from `r length(levels(factor(demo$familyID)))` families spread across four hamlet regions. We excluded test sessions from analysis for the following reasons: refused participation or failure to repeat items presented over headphones even after coaching (N=`r sum(demo$included=="didn\xeat repeat")`), spoke too softly to allow offline coding (N=`r sum(demo$included=="spoke too softly")`), or were 13 years old or older (N=`r sum(demo$Age >12 | demo$included=="too old",na.rm=T)`; we tested these teenagers to put younger children at ease). The remaining `r dim(inc)[1]` children (`r sum(inc$Sex=="G")` girls) were aged from `r min(inc$Age, na.rm=T)` to `r max(inc$Age, na.rm=T)` years (M = `r round(mean(inc$Age, na.rm=T),1)` years, SD = `r round(sd(inc$Age, na.rm=T),1)` years).  In terms of birth order, `r birthorder.tab[[1]]` were first borns, `r birthorder.tab[[2]]` second,  `r birthorder.tab[[3]]` third,  `r birthorder.tab[[4]]` fourth,  `r birthorder.tab[[5]]` fifth, and `r birthorder.tab[[6]]` sixth, with birth order missing for `r sum(is.na(inc$birthOrder))` children. These children were tested in a hamlet far from our research base, and we unfortunately did not ask about birth order before leaving the site. Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years).[^3] We also note that there were `r sum(inc$monolingual=="yes", na.rm=T)` only exposed to Yélî Dnye at home and `r sum(inc$monolingual=="no", na.rm=T)` children exposed to Yélî Dnye plus one or more other languages at home.[^2]

[^2]: Most speakers of Yélî Dnye grow up speaking it monolingually until they begin attending school around the age of 7 years; school instruction is in English. While monolingual Yélî Dnye upbringing is common, multilingual families are not unusual, particularly in the region around the Catholic Mission (the same region in which much of the current data were collected), where there is a higher incidence of married-in mothers from other islands [@brownIPchildrearing]. Children in these multilingual families grow up speaking Yélî Dnye plus English, Tok Pisin, and/or other language(s) from the region.

@@

Taking into account only the first attempts, we derived overall averages across all items. The overall NWR score was M = `r round(mean(overall$Word))`% (SD = `r round(sd(overall$Word))`%), Cohen's d = `r cohensd(mean(overall$Word),sd(overall$Word))`. The phoneme-based normalized Levenshtein distance was  M = `r round(mean(overall$NLD))`% (SD = `r round(sd(overall$NLD))`%), meaning that about a fifth of phonemes were substituted or deleted.

@@

Table \@ref(tab:tab-c) showed that errors involving complex consonants targets were more common than those involving simple consonats (`r 100-mytab["Complex Target","% Corr."]` versus `r 100-mytab["Simple Target","% Corr."]`). Additionally, errors in which a complex consonant was mispronounced as a simple consonant were quite common, whereas those in which a simple consonant was produced as a complex one were vanishingly rare.

@@

### Factors structuring individual variation

@@

Our final exploratory analysis assessed whether variation in scores was structured by factors that vary across individuals, as per our third research question. As shown in Figure \@ref(fig:fig-scores-by-age), there was a greater deal of variance across the tested age range, with significantly higher NWR scores for older children (Spearman's rank correlation, given inequality of variance, rho (`r corage$statistic`) = `r corage$estimate`, p `r pval.display(corage$p.value)`). In contrast, there was no clear association between NWR scores and sex (Welch t (`r tsex$parameter`) = `r tsex$statistic`, p `r pval.display(tsex$p.value)`), birth order  (data missing for `r sum(is.na(inc$birthOrder))` children, rho = (`r round(corbo$statistic,1)`) = `r corbo$estimate`, p `r pval.display(corbo$p.value)`), or maternal education  (rho ( `r round(corme$statistic,1)`) = `r corme$estimate`, p `r pval.display(corme$p.value)`).
